The ANthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2024-01-26 paper.tex
consistently spell the first letter of latin words and phrases in upper case, improve ANNO-FMA-TA2 link description

\newcommand{\aw}{AnthroWorks3D}
\newcommand{\anno}[1]{\href{https://annosaxfdm.de/ontology/#1}{\texttt{#1}}}
\newcommand{\gfo}[1]{\href{https://www.onto-med.de/ontologies/gfo/#1}{\texttt{#1}}}
\newcommand{\fma}[1]{\href{http://purl.org/sig/ont/fma/fma#1}{\texttt{#1}}}
\daurl{anno}{https://annosaxfdm.de/ontology/}
\daurl{gfo}{https://www.onto-med.de/ontologies/gfo\#}
\daurl{fma}{http://purl.org/sig/ont/fma/fma}
\daurl{ta2}{https://ta2viewer.openanatomy.org/?id=}

@@

\section{Introduction}\label{sec:introduction}
Anthropology, as a discipline, explores the full range of human diversity and similarity across cultures, social structures, habitats, and environments.
Anthropological skeletal analysis, a subset of this field, is particularly focused on understanding the human skeletal system.
This study encompasses a multidisciplinary approach, integrating insights from natural, social, ­and cultural sciences.
It's about reading the stories that human remains tell – from sex, age, and health to broader aspects like cultural practices and environmental impacts.
Hence, data from historical and prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, provides a comprehensive understanding of deceased individuals and past populations from different time periods.
This process entails investigating individuals’ identity, health status, aspects of their behaviour, lifestyle, culture and even the circumstances of death~\citep{spurensuche} by determining features (phenotypes) such as sex, age, height and ancestry and other individualizing traits, through the use of anthropological methods.

@@

%\paragraph{Main contributions}
% rest of old text
%ANNO consists of two components: ANNOdc, a domain-core ontology providing core entities such as basic anatomical concepts and classifications, and ANNOds, a domain-specific ontology employed for annotating human skeletal structures.
%
%new text:
Our main contributions are the following:
\begin{enumerate}
\item ANNOdc (\cref{sec:annodc}, \cref{fig:annodc}) is a domain-core (dc) ontology that defines the core entities of the domain.
This includes general anatomical categories (Bone and Tooth), a category for describing their characteristics (Anatomical property), anatomical spatial entities (Anatomical space, surface, line and point) as well as the category Phenotype to model the rules for determining human phenotypes. ANNOdc was embedded in the General Formal Ontology (GFO), i.e. the ANNOdc classes are subclasses of GFO classes.
%
\item ANNOds (\cref{sec:annods}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
These are bones, teeth, their parts and compounds, such as mandible, Mental protuberance or the facial skeleton.
It is also used for modeling their properties and relations, such as the distance between \emph{Mentale dexterum} and \emph{Mentale sinistrum}, needed to derive human phenotypes like sex or height.
ANNOds is embedded in ANNOdc.
In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
%
\item ANNO is integrated into \aw{} (\cref{sec:aw}), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains.
%
\item ANNO is published over multiple channels, see~\cref{tab:publication}.
%
\end{enumerate}

@@

\subsection{Terminologia anatomica (TA)}\label{sec:ta}
The TA~\citep{ta2} is a hierarchy of anatomical structure concepts for the entire human body.
For historical reasons, it uses a terminology consisting of Latin and originally Greek, which was later latinized.
For each anatomical structure, the TA provides the \enquote{preferred}, i.e. standard, Latin term and its English equivalent as well as an individual identification number.
\todo{Marie: bitte mal alle lateinischen Begriffe konsistent groß oder kleinschreiben, momentan ist das total unterschiedlich. Ich traue mich das nicht selbst zu da ich kein Latein kann und nicht weiß, ob da ein Grund dahintersteckt.}
In some cases, Latin and English synonyms are also included, e.g. \emph{malar bone} is an English synonym of the Latin preferred term \emph{Os zygomaticum} and its English equivalent \emph{zygomatic bone}.
The identification number is crucial, as certain landmarks are mostly listed without bone affiliation, resulting in multiple occurrences. %(s. Bone Part)
For instance, the \emph{processus zygomaticus} can be found both at the \emph{os frontale} and the \emph{os temporale}.
Eponyms are terms derived from proper nouns such as the name of a person, place or thing, and they are considered a type of synonyms since they stand for the same anatomical structure.
For example, \emph{ossa digitorum manus / pedis} (engl. \emph{phalanges of the hand}, respectively \emph{foot}) are the small bones of the fingers and toes and are also denoted as \emph{phalanges manus / pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
In addition to the outdated print edition~\citep{ta1998}, an extended second edition (TA2) is available online~\citep{ta2}.
Furthermore, there is a commercially available independent companion print publication with German~\citep{anatomylexicon}, English~\citep{pocketatlas} and Spanish~\citep{taspanish} editions of which at least the German edition is regularly updated.
This book contains textual and visual descriptions, whereby anatomical structures are indicated through lines on the illustrations.

@@

We create manual links between ANNO and the FMA, which we use to connect ANNO with the TA2 by using unofficial links between FMA and TA2 provided by Wikidata.

@@

Anatomical terms, unless historically evolved, typically describe the location, affiliation or function of a concept or structure~\citep{reforminganatomical}.
However, this naming convention is often inconsistent and non-intuitive.
An example, showing the variability in illustrative descriptions are the Tuberculum articulare of the Processus zygomaticus, which in \cite{anatomie} is described as saddle-shaped although it is not called \emph{sella}, the Latin name for saddle.
It is moreover worth pointing out that not every anatomical structure that contributes to an articulation in some way automatically carries the term articular with it, such as the \emph{caput mandibulae}, which connects the Mandibula to the rest of the Cranium.
The Tuber frontale, which serves as an indicator for sex in anthropology, is listed as an eminentia in the TA~\citep{ta2}.
Likewise, the Protuberantia mentale is referred to as Eminentia mentale in a widely-used anthropological method that is also used for sex determination.
As can be seen, the incentive for synonyms is high.
However, the current resources fail to provide comprehensive lists of synonyms in use.
This in turn may lead to misattribution: For example, the FMA entry with the preferred name \emph{External acoustic aperture} and non-English equivalent \emph{Porus acusticus externus} (\aurl{fma}{61301}) lists \emph{external acoustic meatus} (\emph{Meatus acusticus externus} in Latin) as a synonym.
However, it is not the same, rather the Porus is---as its English name indicates---the opening to the meatus~\citep{anatomylexicon}.
%Furthermore, terms oftentimes lack clear affiliation to the bone to which they belong.
%When considering the whole skeleton, these terms consequently occur multiple times, causing ambiguity.
Occasionally, there are discrepancies and inconsistencies found in the definitions of individual anatomical structures.
For instance, the position of the \emph{Corpus ossis pubis} varies across different sources~\citep{anatomylexicon,prometheus,allgemeineanatomie,dualereiheanatomie,anatomiedesmenschen}.

@@

\item ANNOds (\cref{sec:annods}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
These are bones, teeth, their parts and compounds, such as \emph{mandible}, \emph{Mental protuberance} or the facial skeleton.
It is also used for modeling their properties and relations, such as the distance between \emph{Mentale dexterum} and \emph{Mentale sinistrum}, needed to derive human phenotypes like sex or height.
ANNOds is embedded in ANNOdc.
\end{enumerate}
\fi
\begin{figure}[h]
\includegraphics[width=0.6\textwidth]{img/gfo.pdf}
\caption{Integration of ANNO with the top-level GFO ontology.}\label{fig:gfo}
\end{figure}

@@

%\paragraph{Terminology for descriptive anatomy}
\begin{figure}
\includegraphics[width=\textwidth]{img/axisplanelocation.png}
\caption{Left: Anatomical Planes and Axes:
1 saggital plane,
2 midsaggital plane,
3 frontal plane,
4 transverse or horizontal plane,
5 saggital axis,
6 transverse axis,
7 longitudinal or vertical axis.
a: Planum sagittale (sagittal plane), includes sagittal and longitudinal axes; the midsagittal plane passes through the midline of the body,
b: Planum transversale (Transversal plane), includes transverse and sagittal axes,
c: Planum frontale (frontal plane): includes longitudinal and transverse axes.
Right: Anatomical relative location: directional terms used in anatomy and anthropology.
Modified figure based on \cite{allgemeineanatomieen,schemamann}.}
\label{fig:axisplane}
\end{figure}
%\FloatBarrier
The position of all anatomical entities is described in a standardized fashion using conceptual axes (a kind of anatomical line), planes (a kind of anatomical surface) as well as directions (relative anatomical locations).
Directions are relative to the body based on the standard anatomical position (\emph{positio anatomica} \aurl{ta2}{72}, \aurl{fma}{23132}).
It refers to standing upright, facing forward, palms and toes pointing forward, thumbs to the side, and observed from the perspective of the respective individual. %Quelle kommt noch
%
Although an infinite number of axes and planes can be defined within the human body, generally three principal axes and planes are used~\citep{prometheus}.
They are mutually perpendicular and provide the three spatial coordinates.
The three principle axes, shown in \cref{fig:axisplane} are:
\begin{itemize}
\item The longitudinal or vertical axis runs in the superior-inferior direction in an upright position, perpendicular to the ground.
It intersects with the frontal and sagittal planes.
\item The sagittal axis runs in the ventral-dorsal direction, from the front to the back surface of the body and vice versa.
It intersects with the sagittal and transverse planes.
\item The transverse or horizontal axis extends from left to right, intersecting with the frontal and transverse planes.
\end{itemize}

@@

An anatomical line is a conceptual, immaterial line on a bone passing between at least two measurement points that, in the form of distances, circumferences or angles is used to collect measurements representing aspects of a bone's dimensions.
An anatomical line is a boundary (one-dimensional) of an anatomical surface.
Anatomical lines can connect or pass through anatomical entities (e.g., an edge between two or an angle between three anatomical entities).
The length of the line or the angle degree can be measured and used in functions to infer individual phenotypes.% [Verweis zu Phenotype section].
For example, the anatomical line \anno{ZyDexterumZySinistrum} between the \emph{Zygion Dexterum} and the \emph{Zygion Sinistrum} is used in the sex determination function for the phenotype \anno{SexGilesElliot196319}~\citep{sexgileselliot1963}.

@@

\paragraph{Resolving incongruities}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{img/corpus_ossis_pubis.png}
\caption{The \emph{Corpus ossis pubis} (pubic body) forms the proximal part of the \emph{Os pubis} and includes the anterior part of the \emph{Acetabulum} as well as the \emph{Eminentia iliopubica}, which is a small elevation on the \emph{Corpus ossis pubis}.
According to the \emph{TA} and others~\citep{ta2,schemamann,anatomylexicon2007,drake2019gray,anatomylexicon,prometheus2,anatomie}, this landmark is located in the region of the \emph{Facies symphysialis}, which is defined here as the \emph{Pars symphysialis}.
}\label{fig:corpusossispubis}
\end{figure}

The domain experts reconciled conflicting literature to arrive at a consistent and logical result, for example for the \emph{Corpus ossis pubis}, see \cref{fig:corpusossispubis}:
Since the Corpi of the \emph{Os ischii} and \emph{Os ilii} are located in the area of the \emph{Acetabulum},
it is logical to place the \emph{Corpus ossis pubis} in this region as well~\citep{graysanatomy,waldeyer,allgemeineanatomie,datacollection2,romanianmandible}, thereby forming the \emph{Acetabulum} from all Corpi (\emph{Corpus ossis ilii}, \emph{Corpus ossis ischii}, and \emph{Corpus ossis pubis}).
The \emph{Pars symphysialis ossis pubis} has ventral-medially the \emph{Facies symphysialis}, which connects the paired \emph{Ossa pubes} of the pelvic bones.
In the \emph{Pars symphysialis}, the \emph{Ramus inferior ossis pubis} and \emph{Ramus superior ossis pubis} merge into each other.
This redefinition closes the resulting gap in the definition of the area between the Ramus \emph{inferior / superior} ossis pubis.
%
ANNO also corrects the commonly missing differentiation between \emph{Cranium} and \emph{Mandibula} in English literature, with no Latin equivalent for the overall term \enquote{skull}.
Instead, \anno{Cranium} is established as a synonym for the skull~\citep{allgemeineanatomieen,prometheus2}.
\paragraph{}
Further contributions of ANNOds include:
\begin{itemize}
\item Consistent Latin declension (including in plural).
\item Transparency through detailed source references; for measurement distances, origin works and reference works are indicated by dividing into primary and core sources.
\item Dissected group structures.
\item Redefinitions such as joint surfaces, which are often not named in anatomy literature but are anthropologically relevant, and osteometric measurement points outside the skull, which are not established in the literature.
\end{itemize}

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{img/aw3d.png}
%\caption{3D model of a Cranium (skull) inserted into a placeholder skeleton in \aw{} depicting examples of various ANNOdc classes.}
\caption{
Collage of different views in \aw{} of a 3D model of a \emph{cranium} (skull) inserted into a placeholder skeleton (source:~\cite{aw3dcidoc}).
The cranium is a bone compound consisting of 29 bone elements.
One of them, the \emph{mandibula} (mandible), is flexibly connected to the rest of the skull by the mandibular joint.
In the top right, various bone parts on the Mandibula are visually emphasized, with part 14 representing the angulus mandibulae (mandibular angle).
In the top left, the \emph{dens caninus maxillaris dexter} (right maxillary secondary canine tooth), a tooth structure, is highlighted within a circle.
The tooth cusp is a part of the tooth.
It refers to the projection that divides the occlusal surface of most of the teeth, with the exception of the incisors.
At the bottom right, anatomical points (osteometric landmarks) of the right lateral side of the \emph{mandibula} are depicted.
The \emph{gonion dextrum} (right gonion; \emph{gonion dextrum}, abbreviated as \emph{go dextrum}, depicted alternative spelling of dextrum: dexterum) forms part of an anatomical line measuring the \emph{angulus mandibulae},
as shown in the bottom left.
Skeletal material courtesy of Dr. Birgit Grosskopf, Department of Historical Anthropology and Human Ecology, University of Göttingen.
}
\label{fig:aw}
\end{figure}

>>>>

\newcommand{\aw}{AnthroWorks3D}
\newcommand{\anno}[1]{\href{https://annosaxfdm.de/ontology/#1}{\texttt{#1}}}
\newcommand{\gfo}[1]{\href{https://www.onto-med.de/ontologies/gfo/#1}{\texttt{#1}}}
\newcommand{\fma}[1]{\href{http://purl.org/sig/ont/fma/fma#1}{\texttt{#1}}}
\newcommand{\latin}[1]{\emph{#1}}
\daurl{anno}{https://annosaxfdm.de/ontology/}
\daurl{gfo}{https://www.onto-med.de/ontologies/gfo\#}
\daurl{fma}{http://purl.org/sig/ont/fma/fma}
\daurl{ta2}{https://ta2viewer.openanatomy.org/?id=}

@@

\section{Introduction}\label{sec:introduction}
Anthropology, as a discipline, explores the full range of human diversity and similarity across cultures, social structures, habitats, and environments.
Anthropological skeletal analysis, a subset of this field, is particularly focused on understanding the human skeletal system.
This study encompasses a multidisciplinary approach, integrating insights from natural, social, ­and cultural sciences.
It's about reading the stories that human remains tell---from sex, age, and health to broader aspects like cultural practices and environmental impacts.
Hence, data from historical and prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, provides a comprehensive understanding of deceased individuals and past populations from different time periods.
This process entails investigating individuals' identity, health status, aspects of their behaviour, lifestyle, culture and even the circumstances of death~\citep{spurensuche} by determining features (phenotypes) such as sex, age, height and ancestry and other individualizing traits, through the use of anthropological methods.

@@

%\paragraph{Main contributions}
% rest of old text
%ANNO consists of two components: ANNOdc, a domain-core ontology providing core entities such as basic anatomical concepts and classifications, and ANNOds, a domain-specific ontology employed for annotating human skeletal structures.
%
%new text:
Our main contributions are the following:
\begin{enumerate}
\item ANNOdc (\cref{sec:annodc}, \cref{fig:annodc}) is a domain-core (dc) ontology that defines the core entities of the domain.
This includes general anatomical categories (Bone and Tooth), a category for describing their characteristics (Anatomical property), anatomical spatial entities (Anatomical space, surface, line and point) as well as the category Phenotype to model the rules for determining human phenotypes. ANNOdc was embedded in the General Formal Ontology (GFO), i.e. the ANNOdc classes are subclasses of GFO classes.
%
\item ANNOds (\cref{sec:annods}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
These are bones, teeth, their parts and compounds, such as mandible, Mental protuberance or the facial skeleton.
It is also used for modeling their properties and relations, such as the distance between \latin{Mentale dexterum} and \latin{Mentale sinistrum}, needed to derive human phenotypes like sex or height.
ANNOds is embedded in ANNOdc.
In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
%
\item ANNO is integrated into \aw{} (\cref{sec:aw}), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains.
%
\item ANNO is published over multiple channels, see~\cref{tab:publication}.
%
\end{enumerate}

@@

\subsection{Terminologia anatomica (TA)}\label{sec:ta}
The TA~\citep{ta2} is a hierarchy of anatomical structure concepts for the entire human body.
For historical reasons, it uses a terminology consisting of Latin and originally Greek, which was later latinized.
For each anatomical structure, the TA provides the \enquote{preferred}, i.e. standard, Latin term and its English equivalent as well as an individual identification number.
In some cases, Latin and English synonyms are also included, e.g. \emph{malar bone} is an English synonym of the Latin preferred term \latin{Os zygomaticum} and its English equivalent \emph{zygomatic bone}.
The identification number is crucial, as certain landmarks are mostly listed without bone affiliation, resulting in multiple occurrences. %(s. Bone Part)
For instance, the \latin{Processus zygomaticus} can be found both at the \latin{Os frontale} and the \latin{Os temporale}.
Eponyms are terms derived from proper nouns such as the name of a person, place or thing, and they are considered a type of synonyms since they stand for the same anatomical structure.
For example, \latin{Ossa digitorum manus / pedis} (engl. \emph{phalanges of the hand}, respectively \emph{foot}) are the small bones of the fingers and toes and are also denoted as \latin{Phalanges manus / pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
In addition to the outdated print edition~\citep{ta1998}, an extended second edition (TA2) is available online~\citep{ta2}.
Furthermore, there is a commercially available independent companion print publication with German~\citep{anatomylexicon}, English~\citep{pocketatlas} and Spanish~\citep{taspanish} editions of which at least the German edition is regularly updated.
This book contains textual and visual descriptions, whereby anatomical structures are indicated through lines on the illustrations.

@@

We added links from ANNO to the FMA manually, since it required decions from domain experts.
Subsequently, we used those links along with unofficial FMA-TA2 links from Wikidata to automatically generate links between ANNO and the TA2.

@@

Anatomical terms, unless historically evolved, typically describe the location, affiliation or function of a concept or structure~\citep{reforminganatomical}.
However, this naming convention is often inconsistent and non-intuitive.
An example, showing the variability in illustrative descriptions are the Tuberculum articulare of the Processus zygomaticus, which in \cite{anatomie} is described as saddle-shaped although it is not called \latin{Sella}, the Latin name for saddle.
It is moreover worth pointing out that not every anatomical structure that contributes to an articulation in some way automatically carries the term articular with it, such as the \latin{Caput mandibulae}, which connects the Mandibula to the rest of the Cranium.
The Tuber frontale, which serves as an indicator for sex in anthropology, is listed as an eminentia in the TA~\citep{ta2}.
Likewise, the Protuberantia mentale is referred to as Eminentia mentale in a widely-used anthropological method that is also used for sex determination.
As can be seen, the incentive for synonyms is high.
However, the current resources fail to provide comprehensive lists of synonyms in use.
This in turn may lead to misattribution: For example, the FMA entry with the preferred name \emph{External acoustic aperture} and non-English equivalent \latin{Porus acusticus externus} (\aurl{fma}{61301}) lists \emph{external acoustic meatus} (\latin{Meatus acusticus externus} in Latin) as a synonym.
However, it is not the same, rather the Porus is---as its English name indicates---the opening to the meatus~\citep{anatomylexicon}.
%Furthermore, terms oftentimes lack clear affiliation to the bone to which they belong.
%When considering the whole skeleton, these terms consequently occur multiple times, causing ambiguity.
Occasionally, there are discrepancies and inconsistencies found in the definitions of individual anatomical structures.
For instance, the position of the \latin{Corpus ossis pubis} varies across different sources~\citep{anatomylexicon,prometheus,allgemeineanatomie,dualereiheanatomie,anatomiedesmenschen}.

@@

\item ANNOds (\cref{sec:annods}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
These are bones, teeth, their parts and compounds, such as \emph{mandible}, \emph{Mental protuberance} or the facial skeleton.
It is also used for modeling their properties and relations, such as the distance between \latin{Mentale dexterum} and \latin{Mentale sinistrum}, needed to derive human phenotypes like sex or height.
ANNOds is embedded in ANNOdc.
\end{enumerate}
\fi
\begin{figure}[h]
\includegraphics[width=0.6\textwidth]{img/gfo.pdf}
\caption{Integration of ANNO with the top-level GFO ontology.}\label{fig:gfo}
\end{figure}

@@

%\paragraph{Terminology for descriptive anatomy}
\begin{figure}
\includegraphics[width=\textwidth]{img/axisplanelocation.png}
\caption{Left: Anatomical Planes and Axes:
1 saggital plane,
2 midsaggital plane,
3 frontal plane,
4 transverse or horizontal plane,
5 saggital axis,
6 transverse axis,
7 longitudinal or vertical axis.
a: Planum sagittale (sagittal plane), includes sagittal and longitudinal axes; the midsagittal plane passes through the midline of the body,
b: Planum transversale (Transversal plane), includes transverse and sagittal axes,
c: Planum frontale (frontal plane): includes longitudinal and transverse axes.
Right: Anatomical relative location: directional terms used in anatomy and anthropology.
Modified figure based on \cite{allgemeineanatomieen,schemamann}.}
\label{fig:axisplane}
\end{figure}
%\FloatBarrier
The position of all anatomical entities is described in a standardized fashion using conceptual axes (a kind of anatomical line), planes (a kind of anatomical surface) as well as directions (relative anatomical locations).
Directions are relative to the body based on the standard anatomical position (\latin{Positio anatomica} \aurl{ta2}{72}, \aurl{fma}{23132}).
It refers to standing upright, facing forward, palms and toes pointing forward, thumbs to the side, and observed from the perspective of the respective individual. %Quelle kommt noch
%
Although an infinite number of axes and planes can be defined within the human body, generally three principal axes and planes are used~\citep{prometheus}.
They are mutually perpendicular and provide the three spatial coordinates.
The three principle axes, shown in \cref{fig:axisplane} are:
\begin{itemize}
\item The longitudinal or vertical axis runs in the superior-inferior direction in an upright position, perpendicular to the ground.
It intersects with the frontal and sagittal planes.
\item The sagittal axis runs in the ventral-dorsal direction, from the front to the back surface of the body and vice versa.
It intersects with the sagittal and transverse planes.
\item The transverse or horizontal axis extends from left to right, intersecting with the frontal and transverse planes.
\end{itemize}

@@

An anatomical line is a conceptual, immaterial line on a bone passing between at least two measurement points that, in the form of distances, circumferences or angles is used to collect measurements representing aspects of a bone's dimensions.
An anatomical line is a boundary (one-dimensional) of an anatomical surface.
Anatomical lines can connect or pass through anatomical entities (e.g., an edge between two or an angle between three anatomical entities).
The length of the line or the angle degree can be measured and used in functions to infer individual phenotypes.% [Verweis zu Phenotype section].
For example, the anatomical line \anno{ZyDexterumZySinistrum} between the \latin{Zygion Dexterum} and the \latin{Zygion Sinistrum} is used in the sex determination function for the phenotype \anno{SexGilesElliot196319}~\citep{sexgileselliot1963}.

@@

\paragraph{Resolving incongruities}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{img/corpus_ossis_pubis.png}
\caption{The \latin{Corpus ossis pubis} (pubic body) forms the proximal part of the \latin{Os pubis} and includes the anterior part of the \latin{Acetabulum} as well as the \latin{Eminentia iliopubica}, which is a small elevation on the \latin{Corpus ossis pubis}.
According to the \emph{TA} and others~\citep{ta2,schemamann,anatomylexicon2007,drake2019gray,anatomylexicon,prometheus2,anatomie}, this landmark is located in the region of the \latin{Facies symphysialis}, which is defined here as the \latin{Pars symphysialis}.
}\label{fig:corpusossispubis}
\end{figure}

The domain experts reconciled conflicting literature to arrive at a consistent and logical result, for example for the \latin{Corpus ossis pubis}, see \cref{fig:corpusossispubis}:
Since the Corpi of the \latin{Os ischii} and \latin{Os ilii} are located in the area of the \latin{Acetabulum},
it is logical to place the \latin{Corpus ossis pubis} in this region as well~\citep{graysanatomy,waldeyer,allgemeineanatomie,datacollection2,romanianmandible}, thereby forming the \latin{Acetabulum} from all Corpi (\latin{Corpus ossis ilii}, \latin{Corpus ossis ischii}, and \latin{Corpus ossis pubis}).
The \latin{Pars symphysialis ossis pubis} has ventral-medially the \latin{Facies symphysialis}, which connects the paired \latin{Ossa pubes} of the pelvic bones.
In the \latin{Pars symphysialis}, the \latin{Ramus inferior ossis pubis} and \latin{Ramus superior ossis pubis} merge into each other.
This redefinition closes the resulting gap in the definition of the area between the Ramus \emph{inferior / superior} ossis pubis.
%
ANNO also corrects the commonly missing differentiation between \emph{Cranium} and \emph{Mandibula} in English literature, with no Latin equivalent for the overall term \enquote{skull}.
Instead, \anno{Cranium} is established as a synonym for the skull~\citep{allgemeineanatomieen,prometheus2}.
\paragraph{}
Further contributions of ANNOds include:
\begin{itemize}
\item Consistent Latin declension (including in plural).
\item Transparency through detailed source references; for measurement distances, origin works and reference works are indicated by dividing into primary and core sources.
\item Dissected group structures.
\item Redefinitions such as joint surfaces, which are often not named in anatomy literature but are anthropologically relevant, and osteometric measurement points outside the skull, which are not established in the literature.
\end{itemize}

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{img/aw3d.png}
%\caption{3D model of a Cranium (skull) inserted into a placeholder skeleton in \aw{} depicting examples of various ANNOdc classes.}
\caption{
Collage of different views in \aw{} of a 3D model of a \emph{cranium} (skull) inserted into a placeholder skeleton (source:~\cite{aw3dcidoc}).
The cranium is a bone compound consisting of 29 bone elements.
One of them, the \emph{mandibula} (mandible), is flexibly connected to the rest of the skull by the mandibular joint.
In the top right, various bone parts on the Mandibula are visually emphasized, with part 14 representing the angulus mandibulae (mandibular angle).
In the top left, the \latin{Dens caninus maxillaris dexter} (right maxillary secondary canine tooth), a tooth structure, is highlighted within a circle.
The tooth cusp is a part of the tooth.
It refers to the projection that divides the occlusal surface of most of the teeth, with the exception of the incisors.
At the bottom right, anatomical points (osteometric landmarks) of the right lateral side of the \emph{mandibula} are depicted.
The \latin{Gonion dextrum} (right gonion, abbreviated as \latin{Go dextrum}, depicted alternative spelling of dextrum: dexterum) forms part of an anatomical line measuring the \latin{Angulus mandibulae},
as shown in the bottom left.
Skeletal material courtesy of Dr. Birgit Grosskopf, Department of Historical Anthropology and Human Ecology, University of Göttingen.
}
\label{fig:aw}
\end{figure}
